---
title: "Spatial trend analysis with Generalized Ridge GAM"
author: "Javier Roca and Celestino Ordóñez"
date: "2026-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(mgcv)
library(gstat)
library(sp)
library(automap)
library(geodist)
library(spdep)
library(scatterplot3d)
library(sf)
library(units)
library(rgl)
library(ggplot2)
library(MASS)
library(microbenchmark)
library(RANN)
library(Matrix)
```

Define the type of distance, the scenario for the simulated data (1 uniform covariance, 2 non-uniform covariance), the maximum distance between points to be considered and the number of neighbors to be used.
```{r}
metric <- readline(prompt = "Write the type of distance (Euclidean or geodetic): ")
scenario <- as.integer(readline(prompt = "Write the number of the scenario to be simulted: "))
range <- as.integer(readline(prompt = "Write the maximum distance between points for the analysis: "))
spam <- as.integer(readline(prompt = "Write the number of neighbors to be considered: "))
n <- as.integer(readline(prompt = "Write the number of observations: "))
```

Function to fit a Generalized ridge GAM model
INPUTS:
- model: a previous ridge GAM model to be generalized
- w: weights matrix for smoothing
- data: dataframe containing the variables
OUTPUT:
- model: generalized ridge GAM
```{r}
gam.magic=function(model, w, data){
 G <- gam(formula(model), fit=FALSE, data=data)
 mgfit <- magic(G$y, G$X, G$sp, G$S, G$off, rank=G$rank, C=G$C, w=w)
 mg.stuff <- magic.post.proc(G$X, mgfit, w)
 model$edf <- mg.stuff$edf
 model$Vp <- mg.stuff$Vb
 model$coefficients <- mgfit$b 
 model
}
```

This function ensures the covariance matrix (Sigma) is positive definite.
INPUT:
- Sigma: the estimated covariance matrix
OUTPUT:
- Sigma: a numerically stable, positive definite version of the input matrix
```{r}
adjust.Sigma=function(Sigma) {
# https://rubenfcasal.github.io/simbook/fact-cov.html
  tol = 1e-06
  es=eigen(Sigma, symmetric = TRUE)
  ev=es$values  # Sigma eigenvalues
  H=es$vectors # Sigma eigenvectors
  problema=(!all(ev >= -tol * abs(ev)))  # return TRUE if at least there is an eigenvalue less than tol
  
  if (problema) {
   ev[ev<tol]=tol  # 
   Sigma=H%*%diag(ev)%*%t(H)  # Sigma is reconstructed using spectral decomposition S = H * diag(V) * t(h)
  }
  Sigma
}
```
Function to generate a GAM model with nearest neighbors distance dependent coefficients. The output is a matrix with all the data
INPUT: 
 - s: dataset with the coordinates of the points
 - y: dependet variable
 - dist: distance matrix between each pair of points
 - spam = number of nearest neighbors to be consider
 OUTPUT:
 - model: GAM model
 - data: dataframe which column contains the following information: point indices, residual, distance between points and coordinates of the points
```{r}
ModelErrors=function(s, y, dist, spam=5) {
  data2=NULL
  n=nrow(s)
  if (is.matrix(dist) == "FALSE") {dist<-as.matrix(dist)}
  for (i0 in 1:n) {
  ii=order(dist[,i0]) [2:(1+spam)]
  data2=rbind(data2, c(dist[ii,i0], y[ii], y[i0],s[i0,1],s[i0,2]))}

  data2=data.frame(data2)
  names(data2)=c(paste0("dist",1:spam), paste0("y",1:spam), "y","s1","s2")
  
  formula=NULL
  for (j in 1:spam) formula[j]=paste0("s(","dist",j,",by=y",j,")")
  
  formula=paste(formula,collapse="+")
  formula=reformulate(formula, response ="y")
  model=gam(formula=formula, data=data2)
  
  list(model=model, data=data2)
}
```
Function to generate smoothing parameters when the distance is Euclidean
INPUTS:
- s: spatial coordinates
- y: output variable
- range: maximum distance to do the calculations
OUTPUT: a list with the data (columns are i,j,s1,s2,dist,y_i,y_j) and a set of smoothing parameters

```{r}
if (metric == "Euclidean"){
var.model_euclidean = function(s, y, dmat, range=range){
    dist <- as.matrix(dmat)
    data=NULL
    n=length(y)
    for (i in 1:n){
      ii=which(dist[i,]<range & dist[i,]>=0)
      naux=length(ii)
      aux=cbind(rep(i,naux),ii,rep(s[i,1],naux),rep(s[i,2],naux),dist[i,ii],
                rep(y[i],naux),y[ii])
    data=rbind(data,aux)
  }

  dataf=data.frame(data)
  names(dataf)=c("i","j","s1","s2","dist","y1","y2")
  dataf=subset(dataf, i<=j)

  data3=dataf[,c("dist","y1","y2")]

  k=20 # number of baselines
  unc <- gam(y2 ~ s(dist,k=k, bs = "cr", by=y1), data = data3)  # model y2 ~dis(y1,y2)*y1
  sm <- smoothCon(s(dist, k=k, bs = "cr", by=y1), data=data3, knots = NULL)[[1]]
  F <- mono.con(sm$xp, up = FALSE)
  G <- list(X = sm$X, C = matrix(0,0,0), sp = unc$sp,
            p = -sm$xp, y = data3$y2, w =rep(1,nrow(data3)) )


  G$Ain <- F$A    # the monotonicity constraint matrix
  G$bin <- F$b    # the monotonicity constraint vector, both from mono.con
  G$S <- sm$S     # the penalty matrix for the cubic spline
  G$off <- 0      # location of offsets in the penalty matrix
  p <- pcls(G)  # fit spline (using s.p. from unconstrained fit)


  list(data=dataf,sm=sm,p=p)
}
}
```

```{r}
if (metric == 'geodetic') {
geo_distance_matrix <- function(n, radius = 6371) {

  u <- runif(n, cos(pi/6), cos(pi/8))  # change if needed
  theta <- runif(n, 0, cos(pi/20)) # change if needed
  phi <- acos(u)
  
  lat <- 90 - phi * 180/pi
  lon <- -(theta * 180/pi - 180)
  
  pt_geo <- data.frame(lon=lon, lat=lat)
  
  x <- sin(phi) * cos(theta)
  y <- sin(phi) * sin(theta)
  z <- cos(phi)
  
  points_df <- data.frame(lon=lon, lat=lat, x=x, y=y, z=z)
  
  # Generate a sf object win an specific geograpahic CRS
  points_sf <- st_as_sf(points_df,
                        coords = c("lon", "lat"),
                        crs = 4326)  # WGS84
  
  # Compute the distance matrix
  dist_matrix <- st_distance(points_sf)   
  
  # Adjust the radius if it is necessary
  dist_matrix <- dist_matrix * (radius / 6371)
  
  return(list(dist_matrix = dist_matrix, points = points_df, points_geo = pt_geo))
}}
```
Generate points on the ellipsoid and compute the matrix of geometric distances 

```{r}
if (metric == "geodetic"){
  set.seed(123)
  R <- 6371
  result <- geo_distance_matrix(n)
  # Scale unit coordinates to the actual Earth radius
  x <- R * result$points$x
  y <- R * result$points$y
  z <- R * result$points$z
  # 3D visualization using rgl
  plot3d(x, y, z, col="blue", size=5, xlim=c(-R, R), ylim=c(-R, R), zlim=c(-R, R),aspect=c(1,1,1))
  # Add a wireframe sphere to represent the Earth's surface
  spheres3d(0,0,0,radius=R, alpha=0.5, color="black", front="lines", back="lines", box=FALSE)
}
```

Run this chunk only if you want to generate a .png file representing the points on the surface of a sphere
```{r}
# run_chunk <- FALSE
# if (run_chunk) {
#   if (metric=="geodetic"){
#   x <- result$points$x
#   y <- result$points$y
#   z <- result$points$z
#   R <- 6371
#   # Initialize a new rgl window for offline rendering
#   open3d()
#   par3d(windowRect = c(0, 0, 900, 900)) # Set fixed resolution (900x900 px)
#   bg3d("white")  # Set background color to white
#   # Setup the 3D scene without axes or bounding boxes
#   plot3d(0, 0, 0,
#          xlim=c(-R, R),
#          ylim=c(-R, R),
#          zlim=c(-R, R),
#          type="n",   # Plot nothing initially
#          aspect=c(1,1,1),  # Maintain spherical proportions
#          axes = FALSE,   
#          box = FALSE) 
#   # Draw the globe as a wireframe (lightgray lines)
#   spheres3d(0, 0, 0, radius=R, alpha=0.5, color='lightgray', front='lines', back='lines')
#   # Add the simulated blue points on the surface
#   points3d(R*x, R*y, R*z, col="blue", size=5)
#   # Remove any remaining bounding box artifacts and set initial view angle
#   clear3d(type = "bbox")
#   view3d(theta = 0, phi = 0)
#   # Capture the current view and save it as a PNG file
#   rgl.snapshot("sphere.png")
#   # Close the RGL device to free up memory
#   close3d()
#   }
# }
```
Function to create smoothing parameters when the distance is geodetic. The difference with var.model_euclidean is mainly in the type of splines. The dependence is bivariate instead of univariate

```{r}
if (metric == 'geodetic'){
var.model_geodetic=function(s, y, dmat, range=range){
    data=NULL
    n=length(y)
    # Create a pairwise dataset of points within the specified 'range'
    for (i in 1:n){
      ii=which(dmat[i,]<range & dmat[i,]>=0)
      naux=length(ii)
      aux=cbind(rep(i,naux),ii,rep(s[i,1],naux),rep(s[i,2],naux),dmat[i,ii],
                rep(y[i],naux),y[ii])
    data=rbind(data,aux)
  }

  data=data.frame(data)
  names(data)=c("i","j","s1","s2","dist","y1","y2")
  data=subset(data, i<=j)

  data3=data[,c("dist","y1","y2")]
  # Initial unconstrained GAM fit to get starting smoothing parameters
  k = 20
  unc <- gam(y2 ~ s(dist, bs = "cr", by = y1, k = k), data = data3)
  sm <- smoothCon(s(dist, bs = "cr", by = y1, k = k), data = data3, knots = NULL)[[1]]
  print(length(sm$xp))
  F <- mono.con(sm$xp, up = FALSE)   
  G <- list(X = sm$X, C = matrix(0,0,0), sp = unc$sp,
            p = -sm$xp, y = data3$y2, w =rep(1,nrow(data3)) )

  G$Ain <- F$A    # the monotonicity constraint matrix
  G$bin <- F$b    # the monotonicity constraint vector, both from mono.con
  G$S <- sm$S     # the penalty matrix for the cubic spline
  G$off <- 0      # location of offsets in the penalty matrix
  p <- pcls(G)  # fit spline (using s.p. from unconstrained fit)
  
  
  list(data=data,sm=sm,p=p)
}}
```
Data simulation
This function generates a covariance matrix depending on the distance.
Later, the errors with this covariance are simulated
Finally it generates a variable y = trend + errors
INPUTS: 
- dist: distance matrix
- rho: parameter to modify the covariance

```{r}
true.cov=function(dist, rho) {exp(-dist/rho)}  # theoretical covariance
rho <- 20 # correlation range parameter: increase rho to increase the covariance
sig=1.5  # standard deviation of the Gaussian distribution to generate errors

if (metric == 'Euclidean'){
  # Generate random coordinates in a 100x100 square
    s <- cbind(s1=runif(n,0,100),s2=runif(n,0,100))
    dmat <- dist(s)
    # Call the function to create build the covariance matrix
    V <- true.cov(dmat, rho)
    V <- adjust.Sigma(V) # Ensure the matrix is positive-definite
    Cv <- chol(V)     # Cholesky decomposition for error simulation

    } else if (metric=="geodetic") {
      result <- geo_distance_matrix(n,6371)
      dmatrix <- result$dist_matrix
      rho <- rho/100
      dmat <- drop_units(dmatrix) # distance in m
      V <- true.cov(dmat, rho)
      V <- adjust.Sigma(V)
      Cv <- chol(V) 
} else {
  stop("metric must be either 'Euclidean' or 'geodetic'")
}

# Scenario 2: Non-uniform spatial covariance (Heterogeneous structure). We only consider the Euclidean approach, but it could be also valid for the geodetic case
if (metric=="Euclidean" && scenario == 2){
  V1 <- V
  rho1 <- 150
  for (i in 151:300) {  # here n should be 300
    for (j in 151:300){
      dmatrix <- as.matrix(dmat)
      d <- dmatrix[i,j]
      v <- 2*true.cov(d, rho1)
      V1[i,j]=v
    }
  }

  diag(V1) <- diag(V1) + 1e-6

  V2 <- adjust.Sigma(V1)

  Cv <- chol(V2)  # t(Cv)%*%Cv=V

  }  #  end if for scenario 2


# Response variable generation (f + e)
if (metric=="Euclidean"){
  w <- solve(t(Cv))  ## Inverse of Cholesky factor as weights
  e <- t(Cv) %*% rnorm(n, 0, sig) # so cov(e) = V * sig^2 ; These are the simulated errors 
# Trend
  f1=5*sin(8*s[,1]/100)  
  f2=3*(s[,2]/100)
  # f2=6*sin(8*s[,2]/100)
  f=f1+f2
  e <- t(Cv) %*% rnorm(n, 0, sig) # so cov(e) = V * sig^2 ; These are the spatially correlated errors

  y <- f + e 

}

if (metric=="geodetic"){
      w <- solve(t(Cv)) 
      e <- mvrnorm(1, mu = rep(0, n), Sigma = sig^2 * Cv)
      result <- geo_distance_matrix(n,6371)
      points_geo <- result$points_geo
      lon <- points_geo$lon
      lat <- points_geo$lat

  #  trend
    lon_rad <- lon * pi/180
    lat_rad <- lat * pi/180
    f <- 15*sin(8*lon_rad) + 8*lat_rad
    
    y <- f+e
}
```
Function to calculate the distance matrix given a threshold for the distances and using k-nearest neighbors algorithm

```{r}
#library(RANN)
library(Matrix)
library(dbscan)

# dmatrix_dbscan <- function(points, d0 = 80) {
# 
# nn <- frNN(points, eps = d0)
# 
# n <- nrow(points)
# 
# i_list <- vector("list", n)
# j_list <- vector("list", n)
# h_list <- vector("list", n)
# 
# for (i in 1:n) {
#   j <- nn$id[[i]]
# 
#   if (length(j) > 0) {
#     si <- matrix(points[i, ], nrow = length(j), ncol = ncol(points), byrow = TRUE)
#     sj <- points[j, , drop = FALSE]
#     h <- sqrt(rowSums((si - sj)^2))
# 
#     i_list[[i]] <- rep(i, length(j))
#     j_list[[i]] <- j
#     h_list[[i]] <- h
#   }
# }
# 
# i <- unlist(i_list)
# j <- unlist(j_list)
# h <- unlist(h_list)
# 
# D_sparse <- sparseMatrix(i = i, j = j, x = h)
# D_sparse <- (D_sparse + t(D_sparse)) / 2
# }
# 
# # d0 = 80
# # distmatrix <- dmatrix_dbscan(s, d0)
# # V <- true.cov(distmatrix, rho)
# # V <- adjust.Sigma(V) # Ensure the matrix is positive-definite
# # Cv <- chol(V)   # Cholesky decomposition for error simulation
# d0 = 40
# res<-microbenchmark(
#       {
#     distmatrix <- dmatrix_dbscan(s, d0)
#     V <- true.cov(distmatrix, rho)
#     V <- adjust.Sigma(V) # Ensure the matrix is positive-definite
#     Cv <- chol(V)
#     },# Cholesky decomposition for error simulation
#     times = 100)
#     summary_res <- summary(res)
#     median_time <- summary_res$median
#     cat("Elapsed time:", median_time, "nanoseconds\n")
```

Initial visualization of the deterministic trend (f) vs the noisy response (y)
```{r}
if (metric=="Euclidean"){
  par(mfrow=c(1,2))
  s1 <- s[,1]
  s2 <- s[,2]
  ii <- order(s1)
  jj <- order(s2)
  # Plot components of the Euclidean trend
  plot(s1[ii], f1[ii], type="l", main = "Trend f1 (X axis)")
  plot(s2[jj], f2[jj], type="l", main = "Trend f2 (Y axis)")
} else {
  ii <- order(lon)
  lon_ordered <- lon[ii]
  par(mfrow=c(1,2))
  plot(lon_ordered, f[ii], type="l", main = "True Trend (f)", xlab = "Longitude")
  plot(lon_ordered, y[ii], type="l", main = "Observed Response (y)", xlab = "Longitude")
  abline(h=0, lty=3)
}
```

Plot errors over points
```{r}
# library(ggplot2)
# datos_plot <- data.frame(s1 = s[,1], s2 = s[,2], error = e)
# 
# ggplot(datos_plot, aes(x = s1, y = s2, color = error)) +
#   geom_point() +
#   scale_color_gradient2(low = "blue", mid = "white", high = "red",
#                         midpoint = 0) +
#   labs(title = "Simulated error for each point",
#        color = "Error") +
#   theme_minimal()
```
Estimate the covariance of the residuals based on the distance matrix

```{r}
if (metric=="geodetic"){
  # Fit the geodetic variance model using the simulated errors 'e'
  A=var.model_geodetic(points_geo, e, dmat, range=range)
  values=seq(0,range-1000, 15000) # change in needed
  newdata=data.frame(dist=values)
  pred=Predict.matrix(A$sm, newdata) %*% A$p
  new0 = data.frame(dist=0)
  pred0 = Predict.matrix(A$sm, new0) %*% A$p
  pred0 <- pmax(pred0, 0)  # Non-negativity constraint
  pred = pred/as.vector(pred0)
  pred <- pmax(pred, 0) # Ensure covariance values are non-negative
  # Plot estimated covariance (black) vs. theoretical covariance (red)
  plot(values,pred,xlab='Distance (m)', ylab='Covariance', ylim=range(0,1), main = "Covariance Estimation (Geodetic)")
  lines(values,true.cov(values, rho), col=2)

}

if (metric=="Euclidean"){
  A=var.model_euclidean(s, e, dmat, range=range)
  values=seq(0,range-1, 1)
  newdata=data.frame(dist=values)
  pred=Predict.matrix(A$sm, newdata) %*% A$p
  new0 = data.frame(dist=0)
  pred0 = Predict.matrix(A$sm, new0) %*% A$p
  pred0 <- pmax(pred0, 0)  # Non-negativity constraint
  pred = pred/as.vector(pred0)
  pred <- pmax(pred, 0) # Non-negativity constraint
  
  

  plot(values, pred, xlab='Distance (m)', ylab='Covariance', ylim=range(0,1), main = "Covariance Estimation (Euclidean)")
  lines(values,true.cov(values, rho), col=2)
  
  
  
  # err = sqrt(mean(pred-true.cov(values, rho))^2)
  # print(err)

}
```
Unified model estimation
Simultaneously estimates the trend and adjusts for neighbor-based residuals
INPUT:
 - formula: GAM regression equation
 - A: output of var.model_...
 - w: weights matrix derived from the Cholesky factor of the covariance 
 - y: the dependent (response) variable vector
OUTPUT:
 - A list containing both the standard ridge GAM (RGAM) and Generalized ridge GAM (GRGAM)
```{r}
UniqueModel <- function(formula, A, w, y) {
  # Prepare the dataset by merging spatial structures with the response variable
  data2=A$data 
  data2$y=y 
  # Step 1: Fit a standard Ridge GAM (assumes independent errors)
  model0=gam(formula=formula, data=data2)  # ridge gam
  # Step 2: Fit the Generalized Ridge GAM (GRGAM) using the 'magic' algorithm
  # This incorporates the non-identity covariance structure through weights 'w'
  model1=gam.magic(model0, w, data=data2) # generalized ridge gam
  # Store both models in a list for comparison and inference
  models <- list(model0 = model0, model1 = model1)
  return(models)
}
```
# GAM Estimation under known covariance to compare Ridge (RGAM) vs Generalized Ridge (GRGAM)

```{r}
# Higher level of smoothing control with k1, k2 parameters.
k1=k2=6
## Fitting models
if (metric=="Euclidean"){
  dataf=data.frame(s,y)
  colnames(dataf) <- c("s1", "s2", "y")
  s1 <- dataf[,1]
  s2 <- dataf[,2]
  # b: Standard Ridge GAM (assumes identity covariance matrix)
  b <- gam(y ~ s(s1,k=k1)+ s(s2,k=k2), data=dataf) 
  # b2: Generalized Ridge GAM (incorporates known weights 'w')
  b2=gam.magic(b, w, data=dataf)
  } else if (metric=="geodetic"){  # metric is geodetic
      lon <- points_geo$lon
      lat <- points_geo$lat

      dataf <- data.frame(lon,lat,y)
      # Use 'sos' (Splines on the Sphere) for geodetic data to handle spherical geometry    
      b <- gam(y ~ s(lon,lat, bs="sos", k=k1), data=dataf)
      b2=gam.magic(b, w, data=dataf)
  } else{
  print("Metric should be Euclidean or geodetic")
}
## Visualizing Effects and Confidence Intervals
if (metric == "Euclidean"){
  ii=order(s[,1])
  jj=order(s[,2])
  par(mfrow=c(2,2))
  
  # Plot RGAM effects vs True Trend (red line)
  plot(b, main = expression("RGAM (" * rho == 20 * ")"), select=1, scale=0,shade=TRUE)
  lines(s[ii,1], f1[ii]-mean(f1), col=2)
  plot(b, main = expression("RGAM (" * rho == 20 * ")"), select=2,scale=0,shade=TRUE)
  lines(s[jj,2], f2[jj]-mean(f2), col=2)
  
  # Plot GRGAM effects vs True Trend (red line)
  plot(b2, main = expression("GRGAM (" * rho == 20 * ")"), select=1,scale=0,shade=TRUE)
  lines(s[ii,1], f1[ii]-mean(f1), col=2)
  plot(b2, main = expression("GRGAM (" * rho == 20 * ")"), select=2,scale=0,shade=TRUE)
  lines(s[jj,2], f2[jj]-mean(f2), col=2)
} else{  # geodetic distance
  ii <- order(lon)
  lon_ord <- lon[ii]
  lat_fixed <- 65 # Reference latitude for 1D effect comparison (change it if you want)
  newdata <- data.frame(
  lon = lon_ord,
  lat = lat_fixed)
  
  # Extract predictions and standard errors for both models
  predrgam <- predict(b, newdata, type="terms", se.fit=TRUE)
  predgrgam <- predict(b2, newdata, type="terms", se.fit=TRUE)

  fitrgam <- predrgam$fit[,1]
  sergam  <- predrgam$se.fit[,1]
  fitgrgam <- predgrgam$fit[,1]
  segrgam  <- predgrgam$se.fit[,1]
  f_centered <- f[ii] - mean(f[ii])

  par(mfrow=c(2,2))
  # 2D Contour Plot for Ridge GAM
  suppressWarnings(
  vis.gam(b, view=c("lon","lat"), plot.type="contour", main="Ridge GAM")
  )
  # 1D Slice Plot at Latitude 65 for Ridge GAM
  plot(lon, fitrgam, type="l",
       xlim = range(lon_ord),
       ylim = range(fitrgam - 2*sergam, fitrgam + 2*sergam),
       xlab = "Longitude",
       ylab = "Effect",
       main = "Cut at latitude = 65º N")
  
  lines(lon_ord, fitrgam + 2*sergam, lty=2) # Upper confidence band
  lines(lon_ord, fitrgam - 2*sergam, lty=2) # Lower confidence band
  lines(lon_ord, f_centered, col="red", lwd=2) # Ground truth
  abline(h=0, lty=3)
  # 2D Contour Plot for Generalized Ridge GAM
  suppressWarnings(
  vis.gam(b2, view=c("lon","lat"), plot.type="contour", main="Generalized ridge GAM")
  )
  # 1D Slice Plot at Latitude 65 for Generalized ridge GAM
  plot(lon, fitgrgam, type="l",
       xlim = range(lon_ord),
       ylim = range(c(fitgrgam - 2*segrgam, fitgrgam + 2*segrgam, f_centered)),
       xlab = "Longitude",
       ylab = "Effect",
       main = "Cut at latitude = 65ºN")
  
  lines(lon, fitgrgam + 2*segrgam, lty=2)
  lines(lon, fitgrgam - 2*segrgam, lty=2)
  lines(lon, f_centered, col="red", lwd=2)
  abline(h=0, lty=3)
  
}

```
Iterative estimation of trend and covariance components
This loop refines the model by updating the weights (w)
BIC can be used to estimate the adequate value of the number of bases k1=k2

```{r}
# Initialize weights as an Identity matrix (Standard Ridge assumption)
k1 = k2 = 6
# w <- diag(rep(1, n))
max_iter <- 10
bic0 = bic1 <- numeric(max_iter)  # Initialize BIC values
models_list <- list()

if (metric=="Euclidean"){
  ## GAM ignoring correlation (RGAM)
  dataf <- data.frame(s,y)
  colnames(dataf) <- c("s1", "s2", "y")
  s1 <- dataf[,1]
  s2 <- dataf[,2]
  b <- gam(y ~ s(s1,k=k1)+ s(s2,k=k2), data=dataf)
  y <- as.vector(y)
  residuals=y-predict(b) # Initial residuals
  
  for (irep in 1:max_iter) {
    # Increase spline complexity (basis functions) in each iteration
    k1=k1+1
    k2=k2+1
   
    print (paste("Iteration k=",k1))
    # Estimate the spatial variance structure from current residuals
    A=var.model_euclidean(s, residuals, dmat, range=range)
   
    # Visualize Estimated vs Theoretical Covariance
    values=seq(0,range,1)
    newdata=data.frame(dist=values)
    pred=Predict.matrix(A$sm, newdata) %*% A$p
    pred <- pmax(pred, 0)  # Non-negativity constraint
    
    par(mfrow=c(1,1))
    plot(values, pred, type = "l", xlab="Distance (m)", ylab="Covariance")
    lines(values, true.cov (values, rho), col=2, lwd = 2)
    
    # Reconstruct the full Covariance Matrix (V)
    dmat <- as.matrix(dmat)

  for ( i in 1:n) {
   newx=data.frame(dist=dmat[i,]) 
    V[i,]=Predict.matrix(A$sm, newx) %*% A$p
    } # A$sm and A$p were generated from the residuals
  
  # Numerical stability: adjust Sigma and compute Cholesky-based weights  
  diag(V) <- diag(V) + 1e-6   
  V=adjust.Sigma(V)
  
  Cv <- chol(V)  # t(Cv)%*%Cv=V
  w <- solve(t(Cv))  
  
  # Compare standard vs generalized effects in the current step
  b <- gam(y ~ s(s1,k=k1)+ s(s2,k=k2), data=dataf)
  b2=gam.magic(b, w, data=dataf)

  # Visualization of trend recovery
  ii=order(s[,1])
  jj=order(s[,2])
  par(mfrow=c(2,2))
  plot(b, main = expression("RGAM (" * rho == 80 * ")"), select=1, scale=0,shade=TRUE)
  lines(s[ii,1], f1[ii]-mean(f1), col=2)
  plot(b, main = expression("RGAM (" * rho == 80 * ")"), select=2,scale=0,shade=TRUE)
  lines(s[jj,2], f2[jj]-mean(f2), col=2)
  
  plot(b2, main = expression("GRGAM (" * rho == 80 * ")"), select=1,scale=0,shade=TRUE)
  lines(s[ii,1], f1[ii]-mean(f1), col=2)
  plot(b2, main = expression("GRGAM (" * rho == 80 * ")"), select=2,scale=0,shade=TRUE)
  lines(s[jj,2], f2[jj]-mean(f2), col=2)

  # Dynamic Formula Construction for the Unified Model
  formula2=NULL
  for (j in 1:spam) formula2[j]=paste0("s(","dist",j,",by=y",j,")")
  formula2=paste(formula2,collapse="+")
  auxi=paste("s(s1,k=",k1,")+ s(s2,k=",k2,")+")
  
  formula2=paste(auxi, formula2)
  
  formula2=reformulate(formula2, response ="y")
  
  M <- ModelErrors(s, residuals, dmat, spam)
  
  models <- UniqueModel(formula2, M, w, y)
  bic0[irep]=BIC(models$model0)
  bic1[irep]=BIC(models$model1)
  residuals_noridge=as.vector(y)-predict(b) 
  residuals <- y -predict(b)

}
  par(mfrow=c(1,1))
  plot(bic1, type = "b", main='BIC evolution (GRGAM)', xlab="Iteration", ylab="BIC")
  #lines(bic1)
  argumin <- which.min(bic1)
  k1 <- k1 - (max_iter - argumin)
  k2 <- k2 - (max_iter - argumin)
}
```

The same operation of the previous chunk but for geodetic distances
```{r}
if (metric=="geodetic"){
## RGAM
  lon <- points_geo$lon
  lat <- points_geo$lat
  dataf <- data.frame(lon, lat, y)
  colnames(dataf) <- c("lon", "lat", "y")
  b <- gam(y ~ s(lon,lat, bs="sos", k=k1), data=dataf)
  residuals=y-predict(b)

  for (irep in 1:max_iter) {
    k1=k1+1
    k2=k2+1
   
    print (paste("k=",k1))
    A=var.model_geodetic(points_geo, residuals, dmat, range=range)
    # 
    values=seq(0,range-1000, 15000)
    newdata=data.frame(dist=values)
    pred=Predict.matrix(A$sm, newdata) %*% A$p
    pred <- pmax(pred, 0)
    par(mfrow=c(1,1))
    plot(values, pred, type = "l", xlab="Distance", ylab="Covariance", main = paste("Covariance Estimation - Iteration", irep))
    lines(values, true.cov (values, rho), col=2)
  
    # Reconstruct the full Geodetic Covariance Matrix (V)  
    for ( i in 1:n) {
      newx=data.frame(dist=dmat[i,]) 
      V[i,]=Predict.matrix(A$sm, newx) %*% A$p
      } 
    
    # Apply numerical stabilization and update weights via Cholesky decomposition
    diag(V) <- diag(V) + 1e-6   
    V=adjust.Sigma(V)
    
    Cv <- chol(V)  # t(Cv)%*%Cv=V
    w <- solve(t(Cv))  
    
    # Re-fit both standard and generalized models with updated weights
    b <- gam(y ~ s(lon,lat, bs="sos", k=k1), data=dataf)
    b2=gam.magic(b, w, data=dataf)
  
    # lon <- result$points_geo$lon
    ii=order(lon)
    lon_ord <- lon[ii]
    lat_fijo <- 65
    
    newdata <- data.frame(
      lon = lon_ord,
      lat = lat_fijo)
    
    predrgam <- predict(b, newdata, type="terms", se.fit=TRUE)
    predgrgam <- predict(b2, newdata, type="terms", se.fit=TRUE)

    fitrgam <- predrgam$fit[,1]
    sergam  <- predrgam$se.fit[,1]
    fitgrgam <- predgrgam$fit[,1]
    segrgam  <- predgrgam$se.fit[,1]
    f_centered <- f[ii] - mean(f[ii])

    
    par(mfrow=c(1,2))
    
    plot(lon, fitrgam, type="l",
       xlim = range(lon),
       ylim = range(fitrgam - 2*sergam, fitrgam + 2*sergam),
       xlab = "Longitude",
       ylab = "Effect",
       main = "Ridge GAM (65ºN)", cex.main = 0.9)


    lines(lon_ord, fitrgam + 2*sergam, lty=2)
    lines(lon_ord, fitrgam - 2*sergam, lty=2)
    lines(lon_ord, f_centered, col="red", lwd=2)  # Ground truth
    abline(h=0, lty=3)

    plot(lon, fitgrgam, type="l",
       xlim = range(lon),
       ylim = range(c(fitgrgam - 2*segrgam, fitgrgam + 2*segrgam, f_centered)),
       xlab = "Longitude",
       ylab = "Effect",
       main = "Generalized ridge GAM (65ºN)",
       cex.main = 0.9)
  
    lines(lon_ord, fitgrgam + 2*segrgam, lty=2)
    lines(lon_ord, fitgrgam - 2*segrgam, lty=2)
    lines(lon_ord, f_centered, col="red", lwd=2)
    abline(h=0, lty=3)

    # Dynamic formula generation for neighbor-based residuals
    formula2=NULL
    for (j in 1:spam) formula2[j]=paste0("s(","dist",j,",by=y",j, ",bs='cr'", ")")
    formula2=paste(formula2,collapse="+")
    auxi = paste("s(lon, lat, bs='sos', k=",k1, ")+")

    formula2=paste(auxi, formula2)
    
    formula2=reformulate(formula2, response ="y")
    
    M <- ModelErrors(points_geo, residuals, dmat, spam)
    
    models <- UniqueModel(formula2, M, w, y)
    bic0[irep]=BIC(models$model0)
    bic1[irep]=BIC(models$model1)
    residuals_noridge=as.vector(y)-predict(b) 
    residuals <- y -predict(b)
  
  }
  
  par(mfrow=c(1,1))
  plot(bic1, type = "b", main='BIC evolution GRGAM', xlab="Iteration", ylab="BIC")
  #lines(bic1)
    argumin <- which.min(bic1)
  k1 <- k1 - (max_iter - argumin)
  k2 <- k2 - (max_iter - argumin)
  
}
```

Modeling Spatial Residuals and Neighborhood Effects
This section evaluates how much of the remaining error or the original response 
can be explained by the values of the nearest neighbors.

```{r}
if (is.vector(y)=="FALSE"){y <- as.vector(y)}

resid=y-predict(b2, newdata=dataf)

if (metric=="Euclidean"){
# Model A: Residuals as a function of neighbors' residuals  
A=ModelErrors(s,resid,dmat, spam)
# Model B: Original response as a function of neighbors' responses
B=ModelErrors(s,y,dmat, spam) 
 } else{  # Geodetic case using spherical coordinates
A=ModelErrors(points_geo,resid,dmat, spam)
B=ModelErrors(points_geo,y,dmat, spam)
 }

# -Analysis of Model A (Residuals Model)
model_A=A$model
summary(model_A)
# Plot the smooth effects of each neighbor's residual
par(mfrow=c(2,2))
plot(model_A, pages=1)
# # Diagnostic plot: Observed Residuals vs Predicted Residuals
par(mfrow=c(1,1))
plot(A$data$y~predict(model_A), xlab = "Predicted Residuals", ylab = "Observed Residuals",
     main = "Residuals Model Fit")
abline(a=0,b=1, col = "red", lwd = 2)

# Analysis of Model B (Full Response Model) 
model_B=B$model
summary(model_B)
# Plot the smooth effects of each neighbor's value
par(mfrow=c(2,2))
plot(model_B, pages=1, , main = "Neighbor Effects on Response")
# Diagnostic plot: Observed Response vs Predicted Response (Local Model)
par(mfrow=c(1,1))
plot(A$data$y~predict(model_B), xlab = "Predicted Response (Local)", ylab = "Observed Response",
     main = "Neighborhood Response Model Fit")
abline(a=0,b=1, col = "red", lwd = 2)

```
--- Unified Model Estimation (RGAM and GRGAM) ---
This block constructs the full model formula combining spatial trend 
and neighborhood-based local correlation effects.
```{r}
if (metric=="Euclidean"){
  # Build the neighborhood formula components: s(dist_j, by = y_j)
  formula=NULL
  for (j in 1:spam) {formula[j]=paste0("s(","dist",j,",by=y",j,")")}
  formula=paste(formula,collapse="+")
  # Combine trend (s1, s2) with neighborhood effects
  formula=paste("s(s1)+s(s2)+", formula)
  formula=reformulate(formula, response ="y")
  
  # Fit both Ridge GAM and Generalized Ridge GAM
  models <- UniqueModel(formula, A, w, y)
  model0 <- models$model0  # Ridge GAM
  model1 <- models$model1  # Generalized Ridge GAM
  # Generate predictions and calculate residuals
  yhat0 <- predict(model0, data=data)
  resd0 <- y -yhat0
  yhat1 <- predict(model1, data=data)
  resd1 <- y - yhat1
  
  # Visual comparison of model fits
  par(mfrow=c(1,2))
  plot(y~yhat0, main = "RGAM Fit (Euclidean)", xlab = "Predicted", ylab = "Observed" )
  abline(a=0,b=1, col = "red", lwd = 2) # Identity line
  plot(y~yhat1, main = "GRGAM Fit (Euclidean)", xlab = "Predicted", ylab = "Observed")
  abline(a=0,b=1, col = "red", lwd = 2)
} else {  # Geodetic case: Build formula using Splines on the Sphere (sos)
  formula=NULL  
  for (j in 1:spam) formula[j]=paste0("s(","dist",j,",by=y",j, ",bs='cr'", ")")
  formula=paste(formula,collapse="+")
  formula=paste("s(lon, lat, bs='sos')+", formula)
  formula=reformulate(formula, response ="y")
  
  models <- UniqueModel(formula, A, w, y)
  model0 <- models$model0  # Ridge GAM
  model1 <- models$model1  # Generalized Ridge gam
  yhat0 <- predict(model0, data=data)
  resd0 <- y -yhat0
  yhat1 <- predict(model1, data=data)
  resd1 <- y - yhat1
  
  par(mfrow=c(1,2))
  plot(y~yhat0, main="RGAM Fit (Geodetic)", xlab = "Predicted", ylab = "Observed")
  abline(a = 0, b = 1, col = "red", lwd = 2)
  plot(y~yhat1, main="GRGAM Fit (Geodetic)", xlab = "Predicted", ylab = "Observed")
  abline(a = 0, b = 1, col = "red", lwd = 2)
}
```
Spatial Autocorrelation Analysis (Moran's I) 
This block tests whether the residuals of the models still exhibit 
spatial patterns or if they have been successfully reduced to white noise.

```{r}
if (metric=="Euclidean"){
  # Define spatial neighbors based on a fixed distance radius (Adjust d2 if needed)
  coords <- cbind(s[,1], s[,2])
  nb <- dnearneigh(coords, d1 = 0, d2 = 60) 
  # Convert neighbor list to a row-standardized weight matrix
  lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
  # Ensure residuals are in vector format for the test
  resi0v <- as.vector(resd0)
  resi1v <- as.vector(resd1)
  # lisa <- localmoran(resiv, lw)
  # summary(lisa)
  # plot(lisa[,1])
  # Perform Moran's I test for both models
  im0 <- moran.test(resi0v, lw) # Standard Ridge GAM
  im1 <- moran.test(resi1v, lw) # Generalized Ridge GAM
  
  cat('Moran Index - Ridge GAM:', im0$estimate[1], 'p-value', im0$p.value,'\n')
  cat('Moran Index - Gen. Ridge GAM:', im1$estimate[1], 'p-value', im1$p.value,'\n')
} else  { # Geodetic case using Longitude and Latitude coordinates
  coords <- cbind(lon, lat)
  # Define neighbor radius for spherical data (Adjust d2 as necessary)
  nb <- dnearneigh(coords, d1 = 0, d2 = 10000) 
  lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
  resi0v <- as.vector(resd0)
  resi1v <- as.vector(resd1)
  # lisa <- localmoran(resiv, lw)
  # summary(lisa)
  # plot(lisa[,1])
  im0 <- moran.test(resi0v, lw)
  im1 <- moran.test(resi1v, lw)
  
  cat('Moran Index - Ridge GAM:', im0$estimate[1], 'p-value', im0$p.value,'\n')
  cat('Moran Index - Generalized Ridge GAM:', im1$estimate[1], 'p-value', im1$p.value,'\n') 
  
}

```
# Spatial correlation analysis via semivariogram
# This block uses classical geostatistics to evaluate the spatial structure 
# of the residuals and verify if any spatial dependency remains.
```{r}
# Create a spatial object
datos_sp <- data.frame(resi0v)

coordinates(datos_sp) <- coords
# Compute the Empirical Variogram
timevariog <- microbenchmark(
  pipeline = {
variograma_empirico <- variogram(datos_sp@data$resi0v~datos_sp@coords[,1]+datos_sp@coords[,2],data=datos_sp)
# Plot the raw empirical variogram
# plot(variograma_empirico, main = "Empirical Variogram of Residuals")
# Automatic Variogram Model Fitting
# Fits the best theoretical model among Spherical (Sph), Exponential (Exp), 
# and Gaussian (Gau) using weighted least squares.
# modelo_variograma_ajustado <- fit.variogram(variograma_empirico, model = vgm(c("Sph", "Exp", "Gau")))
modelo_variograma_ajustado <- fit.variogram(variograma_empirico, model = vgm("Sph"))
},
times = 100
)
median_ms <- summary(timevariog)$median
cat("Median execution time:", median_ms, "ns\n")
# Overlay the fitted theoretical model onto the empirical points
plot(variograma_empirico, modelo_variograma_ajustado, main = "Empirical Variogram and Fitted Model (GAM Residuals)")

```
## Estimate computational time to create a semivariogram
```{r}
# n <- 5000
# coords <- cbind(s1=runif(n,0,100),s2=runif(n,0,100))
# resim <- runif(n,-1,1)
# datos_sp <- data.frame(resim)
# coordinates(datos_sp) <- coords
# datos_sp$coords[1]
# # Compute the Empirical Variogram
# timevariog <- microbenchmark(
#   pipeline = {
# variograma_empirico <- variogram(datos_sp@data$resim~datos_sp@coords[,1]+datos_sp@coords[,2],data=datos_sp)
# modelo_variograma_ajustado <- suppressWarnings(fit.variogram(variograma_empirico, model = vgm("Sph")))
# },
# times = 100
# )
# median_ms <- summary(timevariog)$median
# cat("Median execution time:", median_ms, "ns\n")
# # Overlay the fitted theoretical model onto the empirical points
# plot(variograma_empirico, modelo_variograma_ajustado, main = "Empirical Variogram and Fitted Model (GAM Residuals)")
```

# Spatial Correlation Analysis for Generalized Ridge GAM (GRGAM) 
# This block examines the residuals of the final generalized model to 
# ensure that all spatial dependency has been accounted for.

```{r}
# Create a Spatial Object for GRGAM residuals
datos_sp <- data.frame(resi1v)
coordinates(datos_sp) <- coords
datos_sp$coords[1]
# Compute the Empirical Variogram for GRGAM residuals
variograma_empirico <- variogram(datos_sp@data$resi1v~datos_sp@coords[,1]+datos_sp@coords[,2],data=datos_sp)
# Plot the empirical variogram
plot(variograma_empirico, main = "Empirical Variogram (GRGAM Residuals)")
# Fit a Theoretical Variogram Model
modelo_variograma_ajustado <- fit.variogram(variograma_empirico, model = vgm(c("Sph", "Exp", "Gau")))
# Final Diagnostic Visualization
plot(variograma_empirico, modelo_variograma_ajustado, main = "Empirical Variogram and Fitted Model (GRGAM residuals)")
```

# Final Model Inference 
# Estimating the response variable 'y' by combining spatial coordinates 
# and localized information from the nearest neighbors

```{r}
data2=A$data
if (metric=="geodetic"){
  names(data2)[names(data2)=="s1"] <- "lon"
  names(data2)[names(data2)=="s2"] <- "lat"
}
# Substitute the simulated error with the actual response variable 'y'
data2$y=y  
# Fit the models: model0 (Standard Ridge) and model1 (Generalized with magic algorithm)
model0=gam(formula=formula, data=data2)  
model1=gam.magic(model0, w, data=data2) 

# --- Visualization for Euclidean Metric ---
if (metric=="Euclidean"){
  par(mfrow=c(2,2))
  # Plot Trend Effects (s1 and s2 components)
  plot(model0, main = "RGAM: spatial trend", select=1,scale=0, shade=TRUE)
  lines(s[ii,1], f1[ii]-mean(f1), col=2) # ground truth
  plot(model0, main = "RGAM: Spatial Trend", select=2,scale=0,shade=TRUE)
  lines(s[jj,2], f2[jj]-mean(f2), col=2)
  
  plot(model1, main = "GRGAM: Spatial Trend", select=1,scale=0,shade=TRUE)
  lines(s[ii,1], f1[ii]-mean(f1), col=2)
  plot(model1, main = "GRGAM: Spatial Trend", select=2,scale=0,shade=TRUE)
  lines(s[jj,2], f2[jj]-mean(f2), col=2)
  
  # Plot Neighborhood Effects (Local components)
  par(mfrow=c(2,2))
  plot(model0, main = "RGAM: Neighbor 1", select=3,scale=0,shade=TRUE)
  plot(model0, main = "RGAM: Neighbor 2", select=4,scale=0,shade=TRUE)
  plot(model1, main = "GRGAM: Neighbor 1", select=3,scale=0,shade=TRUE)
  plot(model1, main = "GRGAM: Neighbor 1", select=4,scale=0,shade=TRUE)
  
  summary(model0)
} else {  # geodetic distance
  # Plot Spherical Trend Recovery (SOS splines)
  par(mfrow=c(1,2))
  plot(model0, main = "GRGAM: Geodetic Trend", select=2,scale=0, shade=TRUE)
  lines(lon[ii], f[ii]-mean(f[ii]), col=2)

  plot(model1, main = "GRGAM: Geodetic Trend", select=2,scale=0,shade=TRUE)
  lines(lon[ii], f[ii]-mean(f), col=2)

  
  # Plot Nearest Neighbor effects for Geodetic data
  par(mfrow=c(1,2))
  plot(model0, main = "RGAM: Neighbor Effect", select=3,scale=0,shade=TRUE)
  plot(model1, main = "GRGAM: Neighbor Effect", select=3,scale=0,shade=TRUE)

  
  summary(model1)
}
```

# Graphical comparison of Model Performance
# Comparing the fitted values of RGAM vs GRGAM to evaluate the 
# contribution of global trends and local neighborhood effects

```{r}
if (metric=="Euclidean"){
  par(mfcol=c(2,2))
  # Full Model Comparison (Trend + Neighbors)
  plot(data2$y, predict(model0), main="Complete RGAM", xlab = "Observed", ylab = "Predicted")
  abline(a=0,b=1, col = 2, lwd = 2)
  plot(data2$y, predict(model1), main="Complete GRGAM", xlab = "Observed", ylab = "Predicted")
  abline(a=0,b=1)
  
  # Trend-Only Comparison
  # Neutralize neighborhood effects by setting neighbor distances and values to zero
  newdata=data2; newdata[,1:(2*spam)]=0  
  # Predict using ONLY spatial coordinates (s1, s2)
  pred0=predict(model0, newdata=newdata)
  plot(data2$y, pred0, main="Only trend RGAM", xlab = "Observed", ylab = "Predicted")
  abline(a=0,b=1,col=2, lwd = 2)
  pred1=predict(model1, newdata=newdata)
  plot(data2$y, pred1, main="Only trend GRGAM", , xlab = "Observed", ylab = "Predicted")
  abline(a=0,b=1,col=2, lwd = 2)
} else {  # # Geodetic distance 
  # names(data2)[names(data2)=="s1"] <- "lon"
  # names(data2)[names(data2)=="s2"] <- "lat"
  # Full Model Comparison
  par(mfcol=c(2,2))
  plot(data2$y, predict(model0), main="Complete RGAM (geodetic)", xlab = "Observed", ylab = "Predicted")
  abline(a=0,b=1, col=2, lwd = 2)
  plot(data2$y, predict(model1), main="Complete GRGAM", xlab = "Observed", ylab = "Predicted")
  abline(a=0,b=1, col=2, lwd = 2)
  
  # Trend-Only Comparison
  newdata=data2
  newdata[,1:(2*spam)]=0  # without data from the neighbors
  pred0=predict(model0, newdata=newdata)
  plot(data2$y, pred0, main="Only trend RGAM (geodetic)", xlab = "Observed", ylab = "Predicted")
  abline(a=0,b=1,col=2, lwd = 2)
  pred1=predict(model1, newdata=newdata)
  plot(data2$y, pred1, main="Only trend GRGAM (geodetic", xlab = "Observed", ylab = "Predicted")
  abline(a=0,b=1, col=2, lwd = 2)
}
```
# Local Indicators of spatial sssociation (LISA) for RGAM
# This block identifies specific geographic areas where residuals 
# may still be clustered, providing a localized diagnostic of the model

```{r}
# Extract residuals from the model
pred <- predict(model0, data=data2)
resi <- data2$y - as.vector(pred)  # data2$y is y because we changed it before
if (metric=="Euclidean") {dmax <- 80
} else {dmax <- 15000}
nb <- dnearneigh(coords, d1 = 0, d2 = dmax) # change d2 if distance is Euclidean
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
resiv <- as.vector(resi)
# Compute Local Moran's I (LISA)
# This estimates spatial autocorrelation for each individual point.
lisa <- localmoran(resiv, lw)
summary(lisa)

# Diagnostic Plots
# Plotting the local Moran coefficients helps identify 'hotspots' or 'outliers'
par(mfrow = c(1,1))
plot(lisa[,1], main = "Local Moran's I Coefficients", 
     xlab = "Index", ylab = "Local I", pch = 19, col = "steelblue")
abline(h = 0, lty = 2, col = "red")

# Global Moran's Test for Final Confirmation
moran.test(resi, lw)
```

# Local Indicators of spatial sssociation (LISA) for GRGAM
# This block identifies specific geographic areas where residuals 
# may still be clustered, providing a localized diagnostic of the model

```{r}
# Extract residuals from the model
pred <- predict(model1, data=data2)
resi <- data2$y - as.vector(pred)  # data2$y is y because we changed it before
if (metric=="Euclidean") {dmax <- 80
} else {dmax <- 15000}
nb <- dnearneigh(coords, d1 = 0, d2 = dmax) # change d2 if distance is Euclidean
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
resiv <- as.vector(resi)
# Compute Local Moran's I (LISA)
# This estimates spatial autocorrelation for each individual point.
lisa <- localmoran(resiv, lw)
summary(lisa)

# Diagnostic Plots
# Plotting the local Moran coefficients helps identify 'hotspots' or 'outliers'
par(mfrow = c(1,1))
plot(lisa[,1], main = "Local Moran's I Coefficients", 
     xlab = "Index", ylab = "Local I", pch = 19, col = "steelblue")
abline(h = 0, lty = 2, col = "red")

# Global Moran's Test for Final Confirmation
moran.test(resi, lw)
```

# Final Hybrid Prediction: Trend + Local Correction 
# This block combines the global spatial trend (b2) with the local 
# neighborhood residuals (A$model) to produce the final estimate

```{r}
data2$y <- y # Substitute errors for the response variable
yhat <- predict(b2, newdata=data2) + predict(A$model, newdata=A$data) 
resid_combined <- data2$y - yhat
# Visual Validation
# Predicted vs Observed values
par(mfrow=c(1,1))
plot(data2$y, yhat, main = "Final Hybrid Model Adjustment", 
     xlab = "Observed Values (y)", ylab = "Hybrid Predictions (y_hat)",
     pch = 19, col = rgb(0.2, 0.4, 0.6, 0.5))
abline(a = 0, b = 1, col = "red", lwd = 2)

# Error Quantification (RMSE)
mse <- mean((yhat-data2$y)^2)
rmse <- sqrt(mse)
cat('RMSE:', rmse,'\n')
data_range <- max(y) - min(y)
rmse_pct_range <- (rmse / data_range) * 100
cat("RMSE (%) relative to Data Range:", round(rmse_pct_range, 4), "%\n")

# RMSE relative to the Standard Deviation (Spread)
data_sd <- sd(y)
rmse_pct_sd <- (rmse / data_sd) * 100
cat("RMSE (%) relative to Standard Deviation:", round(rmse_pct_sd, 4), "%\n")

# RMSE relative to the Mean
data_mean <- mean(y)
rmse_pct_mean <- (rmse / data_mean) * 100
cat("RMSE (%) relative to the Mean:", round(rmse_pct_mean, 4))
```

# Final Diagnostic: Moran's I on Hybrid Residuals
# This test confirms if the combined model (Trend + Neighborhood Correction) 
# has successfully neutralized all spatial dependency

```{r}
# Perform the Moran's I test using the previously defined spatial weights (lw)
im_combined <- moran.test(resid_combined , lw)

# Output the results for the final report
cat('--- Final Spatial Diagnostic ---\n')
cat('Observed Moran I:', round(im_combined$estimate[1], 5), '\n')
cat('Expected Moran I:', round(im_combined$estimate[2], 5), '\n')
cat('P-value:', im_combined$p.value, '\n')

# Quick interpretation logic
if (im_combined$p.value > 0.05) {
  cat("Conclusion: No significant spatial autocorrelation remains. The model is well-specified.\n")
} else {
  cat("Conclusion: Warning! Residual spatial patterns detected. Consider increasing model complexity.\n")
}
```

# Function for Ordinary and Universal Kriging Comparison 
# This function applies classical geostatistical interpolation to a test set
# to benchmark the performance of the Hybrid GAM model

```{r}
krigRes <- function(train, test, vgm_model, spam) {
# Step 1: Prepare Data Frames
# Ensure training and testing sets have consistent coordinate and response naming
train_data <- data.frame(s1 = train$s1, s2 = train$s2, value = train$y)
test_data <- data.frame(s1 = test$s1, s2 = test$s2, value = test$y)

#print(head(test_data))

# Convert to Spatial Objects
# The 'gstat' library requires objects of class 'SpatialPointsDataFrame'
coordinates(train_data) <- ~s1 + s2
coordinates(test_data) <- ~s1 + s2
# Ordinary Kriging (OK)

  kr <- krige(
    formula = value ~ 1,
    locations = train_data,
    newdata = test_data,
    model = vgm_model,
    nmax = spam
  )
  

# Universal Kriging (UK)
# Incorporates a first-order polynomial trend (s1 + s2) 
# This is the classical competitor to the GAM-based approach.  
  KrUniv <- krige(
    formula = value ~ s1 + s2, 
    #formula = value ~ s1 + s2 + I(s1^2) + I(s2^2) + I(s1*s2),
    locations = train_data,
    newdata = test_data,
    model = vgm_model, 
    nmax = spam
    )
  

return (KrUniv)
}
```

# --- Monte Carlo Cross-Validation  ---
# This loop evaluates the predictive power of the proposed Ridge-GAM 
# against Universal Kriging (UK) across multiple random data splits.

```{r}
set.seed(132)
nrep <- 30 # Number of random split iterations
RMSERR = RMSE = RMSEKO = RMSEKU = RMSETOT = RMSERRCOMB = numeric(nrep)

df_train_test <- function(datos, porcentaje){
  n_filas_total <- nrow(datos)
  n_train <- round(n_filas_total * porcentaje/100)
  indices_train <- sample(n_filas_total, n_train)
  df_train <- datos[indices_train, ]
  indices_test <- setdiff(1:n_filas_total, indices_train)
  df_test <- datos[indices_test, ]
  return (list(df_train, df_test, indices_train, indices_test))
}

# Step 1: Data Partitioning
# Split into 90% Training and 10% Testing
for (irep in 1:nrep){
  dtraintest <- df_train_test(dataf, 90)
  datostrain <- dtraintest[[1]]  
  datostest <- dtraintest[[2]]
  
  # Extract indices and coordinates
  indextrain <- as.numeric(row.names(datostrain))  # también dtraintest[[3]]
  indextest <- as.numeric(row.names(datostest))   # también dtraintest[[4]]
  strain <- cbind(datostrain[, 's1', drop=FALSE], datostrain[, 's2', drop=FALSE])  # drop=FALSE keeps the names of the rows 
  
# Step 2: Neighborhood Mapping for Test Set
# For each test point, find its nearest neighbors from the training set  
  test_inferencia <- NULL
  indices_train_close_test <- list()  # almacena índices de puntos train próximos a cada punto test
  i <- 1
  for (i0 in indextest) {
    # Find the indices of the 'spam' nearest neighbors in the training pool
    orden_matrix=order(dmat[indextrain, i0]) [1: spam]
    ii <- indextrain[orden_matrix]
    # cat('i0 ',i0,'\n')
    # cat('ii ',ii, '\n')
    test_inferencia <- rbind(test_inferencia, c(dmat[ii,i0], y[ii], y[i0],s[i0,1],s[i0,2]))
    test_inferencia <- data.frame(test_inferencia)
    names(test_inferencia) <- c(paste0("dist",1:spam), paste0("y",1:spam), "y","s1","s2")
    indices_train_close_test[[i]] <- ii
    i <- i+1
  }
  rownames(test_inferencia) <- as.character(indextest)
  
  train_inferencia <- NULL
  indices_train_close_train <- list()  # stores indices of train points near each train point
  i <- 1
  for (i1 in indextrain) {
    # cat('i0',i0,'\n')
    orden_matrix=order(dmat[indextrain,i1]) [2: (1+spam)]  ## relative position in the matrix
    # cat('orden_matrix',orden_matrix, '\n')
    ii <- indextrain[orden_matrix]  ## point index
    # cat('ii',ii, '\n')
    train_inferencia <- rbind(train_inferencia, c(dmat[ii,i1], y[ii], y[i1],s[i1,1],s[i1,2])) ## Index ii is used because dmat, y and s contain all the data (n in our case) 
    train_inferencia <- data.frame(train_inferencia)
    names(train_inferencia) <- c(paste0("dist",1:spam), paste0("y",1:spam), "y","s1","s2")
    indices_train_close_train[[i]] <- ii  # relative order. The first item in the list would correspond to the index of the first element in `indextrain`
    i <- i + 1
  }
  rownames(train_inferencia) <- as.character(indextrain)  # the row names become the indices of the vertices
  ytrain_inferencia <- train_inferencia$y
  ytrain_inferencia <- as.vector(ytrain_inferencia)
  btrain <- gam(ytrain_inferencia ~ s(s1,k=k1) + s(s2,k=k2), data=train_inferencia)
  b2train <- gam.magic(btrain, w[indextrain, indextrain], data=train_inferencia)  ## w has dimensions n×n, so you can use indextrain as an index
  
  residuos_train <-  ytrain_inferencia-predict(b2train)
  rownames(residuos_train) <- indextrain 
  A_train <- ModelErrors(strain, residuos_train, dmat[indextrain,indextrain], spam)  
  
  modelo_train=A_train$model  # model with distances to the nearest neighbours and no trend
  summary(modelo_train)
  # par(mfrow=c(2,2))
  # plot(modelo_train, pages=1)
  
  # par(mfrow=c(1,1))
  # plot(A_train$data$y~predict(modelo_train))
  # abline(a=0,b=1)
  
  datos_residuos_train <- A_train$data
  datos_residuos_train$y <- ytrain_inferencia  # replace errors with values y at the points
  modelo0_train=gam(formula=formula, data=datos_residuos_train)  # a model in which the output is y, but the data points y1, y2, etc. are residuals. It aims to estimate y based on the residuals from the nearest neighbours, adding the trend (s(s1) + s(s2))
  modelo1_train=gam.magic(modelo0_train, w[indextrain, indextrain], data =datos_residuos_train) # generalized ridge GAM
  
  # par(mfrow=c(2,2))
  # plot(modelo0_train, main = "Ignoring correlation", select=1,scale=0, shade=TRUE)
  # points(strain[ii,1], f1[ii]-mean(f1[ii]), col=2)
  # plot(modelo0_train, main = "Ignoring correlation", select=2,scale=0,shade=TRUE)
  # points(strain[jj,2], f2[jj]-mean(f2[jj]), col=2)
  # 
  # plot(modelo1_train, main = "Known correlation", select=1,scale=0,shade=TRUE)
  # lines(strain[ii,1], f1[ii]-mean(f1[ii]), col=2)
  # plot(modelo1_train, main = "Known correlation", select=2,scale=0,shade=TRUE)
  # points(strain[jj,2], f2[jj]-mean(f2[jj]), col=2)
  # 
  # 
  # par(mfrow=c(2,2))
  # plot(modelo0_train, main = "Ignoring correlation", select=3,scale=0,shade=TRUE)
  # plot(modelo0_train, main = "Ignoring correlation", select=4,scale=0,shade=TRUE)
  # plot(modelo1_train, main = "Known correlation", select=3,scale=0,shade=TRUE)
  # plot(modelo1_train, main = "Known correlation", select=4,scale=0,shade=TRUE)
  
  #summary(modelo0_train)
  # summary(modelo1_train)  # modelo ridge
  
  # par(mfcol=c(1,2))
  # plot(datos_residuos_train$y, predict(modelo0_train),main="Modelo Completo");abline(a=0,b=1, col=2)
  #plot(datos2$y, predict(modelo1)); abline(a=0,b=1)
  
  newdata_train=datos_residuos_train; newdata_train[,1:2*spam]=0  # don't provide the neighbours' details
  
  pred_train=predict(modelo0_train, newdata=newdata_train)
  # plot(datos_residuos_train$y, pred_train, main="Only trend");abline(a=0,b=1,col=2)
  
  pred_train <- predict(modelo0_train, data=train_inferencia)
  resi_train <- datos_residuos_train$y - as.vector(pred_train)  # Remember that datos_residuos_train$y is y
  coords_train <- strain

  
  matrix_train_close_test <- do.call(rbind, indices_train_close_test)  # row names are relative
  matrix_train_close_train <- do.call(rbind, indices_train_close_train)  # lrow names are relative
  test_inferencia_residuos <- test_inferencia  # A copy of the test_inferencia in which the values of y must be replaced with residuals
  
  
  for (i in 1:dim(test_inferencia)[1]){
    itest <- indextest[i]  # not used
    for (j in 1:spam){
      itrain <- matrix_train_close_test[i,j]  # neighbour of train a test
      residuo <- residuos_train[as.character(itrain)]  
      test_inferencia_residuos[i,spam+j] <- residuo
    }
  }
  
  ## RMSE UNIFIED RGAM
  pred_test_inferencia <- predict(modelo0_train, newdata=test_inferencia_residuos)
  error_test_inferencia <- pred_test_inferencia - test_inferencia$y
  # plot(pred_test_inferencia, test_inferencia$y)
  # abline(a=0,b=1)
  rmse_inferencia = sqrt(mean(error_test_inferencia^2))
  
  ## RMSE UNIFIED GRGAM
  pred_test_inferencia_RR <- predict(modelo1_train, newdata=test_inferencia_residuos)
  error_test_inferencia_RR <- pred_test_inferencia_RR - test_inferencia$y
  # plot(pred_test_inferencia_RR, test_inferencia$y)
  # abline(a=0,b=1)
  rmse_inferencia_RR = sqrt(mean(error_test_inferencia_RR^2))
  
  ## RMSE TREND + RESIDUALS
  pred_test_inferencia_trend <- predict(b2train, newdata=datostest)
  yhat_model_comb <- pred_test_inferencia_trend + predict(A_train$model, newdata=test_inferencia_residuos)
  err_test_inferencia_RRcom <- as.vector(yhat_model_comb) - test_inferencia$y
  # plot(yhat_model_comb, test_inferencia$y)
  # abline(a=0,b=1)
  rmse_inferencia_RRcom = sqrt(mean(err_test_inferencia_RRcom^2))
 
  RMSE[irep] <- rmse_inferencia
  RMSERR[irep] <- rmse_inferencia_RR
  RMSERRCOMB[irep] <- rmse_inferencia_RRcom
  
  ## Krigeado
  
  ## Create a gstat object
  s1 <- datostrain$s1
  s2 <- datostrain$s2
  coordinates(datostrain) <- ~s1 + s2
  # gkor <- gstat(formula = y ~ 1, data = datostrain)
      # gkun <- gstat(formula = y ~ s1 + s2, data = traint)
  gkun <- gstat(formula = y ~ s1 + s2 , data = datostrain)
      
  # emp_variogram_kor <- variogram(gkor)
  emp_variogram_kun <- variogram(gkun)
      

  vgm_model_fit_un <- autofitVariogram(
        formula = y ~ s1 + s2,
        input_data = datostrain
  )
      

      
  KrU <- krigRes(datostrain, datostest, vgm_model_fit_un$var_model, spam)  # [[2]]
  
  # par(mfcol=c(1,2))
  # 
  # par(mfrow=c(1,2))
  # plot(pred_test_inferencia, test_inferencia$y)
  # abline(a=0,b=1)
  # plot(KrU$var1.pred, test_inferencia$y)
  # abline(a=0,b=1)
  

  RMSEKU[irep] = sqrt(mean((datostest$y-KrU$var1.pred)^2))

}

  
par(mfcol=c(1,1))  
boxplot(cbind(RMSE, RMSERRCOMB, RMSEKU), names = c('RGAM', 'GRGAM', 'UKr'), ylab='RMSE', outline=TRUE)
```
# Application to real data
## Load the file with the data
## https://www.aspexit.com/linear-modelling-of-spatial-data-in-r/

```{r}
data(meuse)
meuse <- st_as_sf(meuse, coords=c('x','y'), remove=FALSE)
summary(meuse)
st_geometry(meuse)=NULL
```
Create a dataframe with columns s1,s2,y

```{r}
datos <- data.frame(s1=meuse$x, s2=meuse$y, y=meuse$lead, dist=meuse$dist)
s <- cbind(datos$s1, datos$s2)
```

# Represent data for Pb
```{r}
library(ggplot2)
library(ggspatial)
# y <- as.vector(datos$y)
# s1 <- datos$s1
# s2 <- datos$s2
ggplot(datos, aes(x = s1, y = s2, color = y)) +
  geom_point() +
  scale_color_gradient2(name = waiver(), low = "blue", mid = "yellow", high = "red",
                        midpoint = 300) +
  labs(title = "",
        x = expression(s[1]),
        y = expression(s[2]),
       color = "Pb (ppm)") +
  theme_minimal() +
    annotation_north_arrow(
    location = "tl",          # esquina top-right
    which_north = "true",     # norte verdadero
    style = north_arrow_fancy_orienteering
  )
```

Build de distance matrix

```{r}
distancias <- dist(s)
dmatrix <- as.matrix(distancias)
```

Residuals covariance

```{r}
rango <- 3000
y <- as.vector(datos$y)
A <- var.model_euclidean(s, y, dmatrix, range=rango)
values <- seq(0,rango-1,1)
newdata <- data.frame(dist=values)
pred <- Predict.matrix(A$sm, newdata) %*% A$p
new0 <- data.frame(dist=0)
pred0 <- Predict.matrix(A$sm, new0) %*% A$p
pred <- pred/as.vector(pred0)
pred <- pmax(pred, 0)

# pdf_path <- paste(dir,'figuras', 'variance_scenario2.pdf',sep='/')
# pdf(pdf_path)
plot(values,pred,xlab='Distance', ylab='Covariance')
# dev.off()
```
RGAM and GRGAM

```{r}
k1 = k2 = 6
n <- nrow(datos)

V <- matrix(0, nrow=n, ncol=n)
for (i in 1:n){
  newx=data.frame(dist=dmatrix[i,])
  V[i,] = Predict.matrix(A$sm, newx) %*% A$p}
V <- V/as.vector(pred0)

diag(V) <- diag(V) + 1e-6
V=(V)

Cv <- chol(V) 

w <- solve(t(Cv)) 

b <- gam(y ~ s(s1, s2, k = k1), data=datos)  # datos contains the data s1, s2, and original
b2=gam.magic(b, w, data=datos)   # W calculated using the Cholesky decomposition of the covariance matrix V

par(mfrow=c(1,2))
# plot(b, main = "GAM", scale=0,shade=TRUE)
#plot(b2, main = "Ridge GAM", scale=0,shade=TRUE)
vis.gam(b, main = "Ridge GAM", view = c("s1", "s2"), plot.type = "contour", color = "topo")
vis.gam(b2, main = "Generalized ridge GAM", view = c("s1", "s2"), plot.type = "contour", color = "topo")

```
RGAM and GRGAM iterating

```{r}
k1 = k2 = 6
b <- gam(y ~ s(s1,k=k1)+ s(s2,k=k2), data=datos)
residuos=y-predict(b)
max_iter <- 10
bic0 = bic1 <- numeric(max_iter)  # Initialize BIC values
models_list <- list()
newx0 = data.frame(dist=0)

# dataf=data.frame(s,y)
# colnames(dataf) <- c("s1", "s2", "y")
# s1 <- dataf[,1]
# s2 <- dataf[,2]
# residuals=y-predict(b)
for (irep in 1:max_iter) {
    # Increase spline complexity (basis functions) in each iteration
    k1=k1+1
    k2=k2+1
   
    print (paste("Iteration k=",k1))
    # Estimate the spatial variance structure from current residuals
    A=var.model_euclidean(s, residuos, dmatrix, range=rango)
   
    # Visualize Estimated vs Theoretical Covariance
    values=seq(0,rango,1)
    newdata=data.frame(dist=values)
    pred=Predict.matrix(A$sm, newdata) %*% A$p
    new0 = data.frame(dist=0)
    pred0 = Predict.matrix(A$sm, new0) %*% A$p
    pred0 <- pmax(pred0, 0)  # Non-negativity constraint
    pred = pred/as.vector(pred0)
    pred <- pmax(pred, 0)  # Non-negativity constraint

    par(mfrow=c(1,1))
    plot(values, pred, type = "l", xlab="Distance (m)", ylab="Covariance")

  for ( i in 1:n) {
   newx=data.frame(dist=dmatrix[i,]) 
    V[i,]=Predict.matrix(A$sm, newx) %*% A$p
    } # A$sm and A$p were generated from the residuals
  
  # Numerical stability: adjust Sigma and compute Cholesky-based weights  
  V <- V/as.vector(pred0)
  diag(V) <- diag(V) + 1e-6   
  V=adjust.Sigma(V)
  
  Cv <- chol(V)  # t(Cv)%*%Cv=V
  w <- solve(t(Cv))  
  
  # Compare standard vs generalized effects in the current step
  b <- gam(y ~ s(s1,k=k1)+ s(s2,k=k2), data=datos)
  b2=gam.magic(b, w, data=datos)

  # Visualization of trend recovery
  par(mfrow=c(2,2))
  plot(b, main = "Ridge GAM", select=1, scale=0,shade=TRUE)
  plot(b, main = "Ridge GAM", select=2,scale=0,shade=TRUE)

  plot(b2, main = "Generalized Ridge GAM", select=1,scale=0,shade=TRUE)
  plot(b2, main = "Generalized Ridge GAM", select=2,scale=0,shade=TRUE)

  # Dynamic Formula Construction for the Unified Model
  formula2=NULL
  for (j in 1:spam) formula2[j]=paste0("s(","dist",j,",by=y",j,")")
  formula2=paste(formula2,collapse="+")
  auxi=paste("s(s1,k=",k1,")+ s(s2,k=",k2,")+")
  
  formula2=paste(auxi, formula2)
  
  formula2=reformulate(formula2, response ="y")
  
  M <- ModelErrors(s, residuos, dmatrix, spam)
  
  models <- UniqueModel(formula2, M, w, y)
  bic0[irep]=BIC(models$model0)
  bic1[irep]=BIC(models$model1)
  residuals_ridge=as.vector(y)-predict(b) 
  residuals_gridge <- as.vector(y) -predict(b2)

}
  par(mfrow=c(1,1))
  plot(bic1, type = "b", main='BIC evolution (GRGAM)', xlab="Iteration", ylab="BIC")
  #lines(bic1)
  argumin <- which.min(bic1)
  k1 <- k1 - (max_iter - argumin)
  k2 <- k2 - (max_iter - argumin)
```

Residuals model
```{r}
residuos=y-predict(b2, newdata=datos)
A=ModelErrors(s,residuos,dmatrix,spam)
B=ModelErrors(s,y,dmatrix,spam)

modelo=A$model
summary(modelo)
par(mfrow=c(2,2))
plot(modelo, pages=1); summary(modelo)

par(mfrow=c(1,1))
plot(A$data$y~predict(modelo))
abline(a=0,b=1)
```

Unified model (RGAM and GRGAM)

```{r}
formula=NULL
for (j in 1:spam) formula[j]=paste0("s(","dist",j,",by=y",j,")")
formula=paste(formula,collapse="+")
formula=paste("s(s1)+s(s2)+", formula)
formula=reformulate(formula, response ="y")

models <- UniqueModel(formula, A, w, y)
model0 <- models$model0  # gam
model1 <- models$model1  # ridge gam
yhat0 <- predict(model0, data=datos)
resd0 <- y -yhat0
yhat1 <- predict(model1, data=datos)
resd1 <- y - yhat1
```

Effects (additive)

```{r}
datos2=A$data
datos2$y=y  # Ojo aquí porque estoy sustituyendo el error por la y del dato
model0=gam(formula=formula, data=datos2)  # modelo en que la salida es la y, pero los datos de y1, y2, etc. son residuos. Quiere estimar la y a partir de los residuos en los vecinos más próximos 
model1=gam.magic(model0, w, data=datos2) # modelo errores suavizado

par(mfrow=c(2,2))
plot(model0, main = "Ignoring correlation", select=1,scale=0, shade=TRUE)
plot(model0, main = "Ignoring correlation", select=2,scale=0,shade=TRUE)

plot(model1, main = "Known correlation", select=1,scale=0,shade=TRUE)
plot(model1, main = "Known correlation", select=2,scale=0,shade=TRUE)



par(mfrow=c(2,2))
plot(model0, main = "Ignoring correlation", select=3,scale=0,shade=TRUE)
plot(model0, main = "Ignoring correlation", select=4,scale=0,shade=TRUE)
plot(model1, main = "Known correlation", select=3,scale=0,shade=TRUE)
plot(model1, main = "Known correlation", select=4,scale=0,shade=TRUE)

summary(model0)
```

Comparative analysis in a test sample

```{r}
set.seed(132)
nrep <- 30  # number of repetitions

RMSERR = RMSE = RMSEKO = RMSEKU =  RMSETOT = RMSERRCOMB = numeric(nrep)

for (irep in 1:nrep){
  
  dtraintest <- df_train_test(datos, 90)  # 90 is the sample size for the training samples
  datostrain <- as.data.frame(dtraintest[[1]])
  datostest <- as.data.frame(dtraintest[[2]])
  
  indextrain <- as.numeric(row.names(datostrain))  # also dtraintest[[3]]
  indextest <- as.numeric(row.names(datostest))   # also dtraintest[[4]]
  strain <- cbind(datostrain[, 's1', drop=FALSE], datostrain[, 's2', drop=FALSE])  # drop=FALSE mantiene los valores de los nombres de las filas
  
  test_inferencia <- NULL
  indices_train_close_test <- list()  # store training point indices nearby each test point
  i <- 1
  for (i0 in indextest) {
    orden_matrix=order(dmatrix[indextrain, i0]) [1: spam]
    ii <- indextrain[orden_matrix]
    # cat('i0 ',i0,'\n')
    # cat('ii ',ii, '\n')
    test_inferencia <- rbind(test_inferencia, c(dmatrix[ii,i0], y[ii], y[i0],s[i0,1],s[i0,2]))
    test_inferencia <- data.frame(test_inferencia)
    names(test_inferencia) <- c(paste0("dist",1:spam), paste0("y",1:spam), "y","s1","s2")
    indices_train_close_test[[i]] <- ii
    i <- i+1
  }
  rownames(test_inferencia) <- as.character(indextest)
  
  train_inferencia <- NULL
  indices_train_close_train <- list()  # almacena índices de puntos train próximos a cada punto train
  i <- 1
  for (i0 in indextrain) {
    # cat('i0',i0,'\n')
    orden_matrix=order(dmatrix[indextrain,i0]) [2: (1+spam)]  ## posición relativa en la matriz
    # cat('orden_matrix',orden_matrix, '\n')
    ii <- indextrain[orden_matrix]  ## índice del punto
    # cat('ii',ii, '\n')
    train_inferencia <- rbind(train_inferencia, c(dmatrix[ii,i0], y[ii], y[i0],s[i0,1],s[i0,2])) ## el índice ii se usa porque dmatrix, y, s, contienen todos los datos (300 en nuestro caso) 
    train_inferencia <- data.frame(train_inferencia)
    names(train_inferencia) <- c(paste0("dist",1:spam), paste0("y",1:spam), "y","s1","s2")
    indices_train_close_train[[i]] <- ii  # orden relativo. El primer elemento de la lista correspondería al índice del primer elemento de indextrain
    i <- i + 1
  }
  rownames(train_inferencia) <- as.character(indextrain)  # names of rows become point indices
  # if (irep==nrep) {
  #   pdf_path <- paste(dir,'figuras', 'points_distribution.pdf',sep='/')
  #   pdf(pdf_path, width = 8, height = 6)
  #   par(mfrow=c(1,1))
  #   plot(s2~s1, data=datos2,  type = "n")
  #   points(s2~s1, data=train_inferencia, col=1,pch=16)
  # # text(x = train_inferencia$s1,
  # #      y = train_inferencia$s2,
  # #      labels = row.names(train_inferencia), # Usa los nombres de las filas como etiquetas
  # #      pos = 4, # Muestra el texto a la derecha del punto
  # #      cex = 0.6, # Reduce un poco el tamaño del texto para que no se superpongan
  # #      col = "blue")
  # #par(mfrow=c(2,2))
  #   points(s2~s1, data=test_inferencia, col=2,pch=16)
  #   # text(x = test_inferencia$s1,
  #   #    y = test_inferencia$s2,
  #   #    labels = row.names(test_inferencia), # Usa los nombres de las filas como etiquetas
  #   #    pos = 4, # Muestra el texto a la derecha del punto
  #   #    cex = 0.6, # Reduce un poco el tamaño del texto para que no se superpongan
  #   #    col = "red")
  #   dev.off()
  # }
 
  ytrain_inferencia <- train_inferencia$y
  ytrain_inferencia <- as.vector(ytrain_inferencia)
  btrain <- gam(ytrain_inferencia ~ s(s1,k=k1) + s(s2,k=k2), data=train_inferencia)
  b2train <- gam.magic(btrain, w[indextrain, indextrain], data=train_inferencia)  ## w tiene dimensión nxn por eso puede usar indextrain como índice
  
  residuos_train <-  ytrain_inferencia-predict(b2train)
  rownames(residuos_train) <- indextrain 
  A_train <- ModelErrors(strain,residuos_train,dmatrix[indextrain,indextrain],spam)  ## Modelo para errores y valores errores en vecino
  
  model_train=A_train$model  # con distancias a vecinos más próximos y sin tendencia
  summary(model_train)
  par(mfrow=c(2,2))
  plot(model_train, pages=1)
  
  par(mfrow=c(1,1))
  plot(A_train$data$y~predict(model_train))
  abline(a=0,b=1)
  
  datos_residuos_train <- A_train$data
  datos_residuos_train$y <- ytrain_inferencia  # sustituyo errores por valor y en los puntos
  model0_train=gam(formula=formula, data=datos_residuos_train)  # modelo en que la salida es la y, pero los datos de y1, y2, etc. son residuos. Quiere estimar la y a partir de los residuos en los vecinos más próximos, sumando la tendencia (s(s1) + s(s2))
  model1_train=gam.magic(model0_train, w[indextrain, indextrain], data=datos_residuos_train) # modelo errores suavizado
  
  par(mfrow=c(2,2))
  plot(model0_train, main = "RGAM", select=1,scale=0, shade=TRUE)
  plot(model0_train, main = "RGAM", select=2,scale=0,shade=TRUE)

  plot(model1_train, main = "GRGAM", select=1,scale=0,shade=TRUE)
  plot(model1_train, main = "GRGAM", select=2,scale=0,shade=TRUE)

  
  par(mfrow=c(2,2))
  plot(model0_train, main = "RGAM", select=3,scale=0,shade=TRUE)
  plot(model0_train, main = "RGAM", select=4,scale=0,shade=TRUE)
  plot(model1_train, main = "GRGAM", select=3,scale=0,shade=TRUE)
  plot(model1_train, main = "GRGAM", select=4,scale=0,shade=TRUE)
  
  #summary(modelo0_train)
  summary(model1_train)  # GRGAM
  
  par(mfcol=c(1,2))
  plot(datos_residuos_train$y, predict(model0_train),main="Modelo Completo");abline(a=0,b=1, col=2)
  #plot(datos2$y, predict(modelo1)); abline(a=0,b=1)
  
  newdata_train=datos_residuos_train; newdata_train[,1:2*spam]=0  # no le doy datos de los vecinos
  
  pred_train=predict(model0_train, newdata=newdata_train)
  plot(datos_residuos_train$y, pred_train, main="Sólo tendencia");abline(a=0,b=1,col=2)
  
  pred_train <- predict(model0_train, data=train_inferencia)
  resi_train <- datos_residuos_train$y - as.vector(pred_train)  # recuerda que datos_residuos_train$y es y
  coords_train <- strain
  # nb <- dnearneigh(coords_train, d1 = 0, d2 = 40)
  # lw <- nb2listw(nb, style = "W")
  # resi_train_vector <- as.vector(resi_train)
  # lisa <- localmoran(resi_train_vector, lw)
  # summary(lisa)
  # plot(lisa[,1])
  # moran.test(resi_train_vector, lw)
  
  matrix_train_close_test <- do.call(rbind, indices_train_close_test)  # los nombres de filas son relativos
  #rownames(matrix_train_close_test) <- indextest
  matrix_train_close_train <- do.call(rbind, indices_train_close_train)  # los nombres de filas son relativos
  #rownames(matrix_train_close_train) <- indextrain
  test_inferencia_residuos <- test_inferencia  # copia de test_inferencia en la que hay que sustituir valores de yi por residuos
  
  
  for (i in 1:dim(test_inferencia)[1]){
    itest <- indextest[i]  # no lo uso
    for (j in 1:spam){
      itrain <- matrix_train_close_test[i,j]  # vecino de train a test
      residuo <- residuos_train[as.character(itrain)]  
      test_inferencia_residuos[i,spam+j] <- residuo
    }
  }
  
  ## RMSE MODELO UNICO
  pred_test_inferencia <- predict(model0_train, newdata=test_inferencia_residuos)
  error_test_inferencia <- pred_test_inferencia - test_inferencia$y
  plot(pred_test_inferencia, test_inferencia$y)
  abline(a=0,b=1)
  rmse_inferencia = sqrt(mean(error_test_inferencia^2))
  
  ## RMSE MODELO ÚNICO RIDGE
  pred_test_inferencia_RR <- predict(model1_train, newdata=test_inferencia_residuos)
  error_test_inferencia_RR <- pred_test_inferencia_RR - test_inferencia$y
  plot(pred_test_inferencia_RR, test_inferencia$y)
  abline(a=0,b=1)
  rmse_inferencia_RR = sqrt(mean(error_test_inferencia_RR^2))
  
  ## RMSE MODELO TENDENCIA MÁS RESIDUOS
  pred_test_inferencia_trend <- predict(b2train, newdata=datostest)
  yhat_model_comb <- pred_test_inferencia_trend + predict(A_train$model, newdata=test_inferencia_residuos)
  err_test_inferencia_RRcom <- as.vector(yhat_model_comb) - test_inferencia$y
  plot(yhat_model_comb, test_inferencia$y)
  abline(a=0,b=1)
  rmse_inferencia_RRcom = sqrt(mean(err_test_inferencia_RRcom^2))
  
  # cat('RMSE', rmse_inferencia,'\n')
  # rango_reales <- max(test_inferencia$y) - min(test_inferencia$y)
  # rmse_porcentaje_rango <- (rmse_inferencia / rango_reales) * 100
  # cat("RMSE (%) basado en el rango de los valores reales:", rmse_porcentaje_rango, "%\n")
  # 
  # sd_reales <- sd(test_inferencia$y)
  # rmse_porcentaje_sd <- (rmse_inferencia / sd_reales) * 100
  # cat("RMSE (%) basado en la desviación estándar de los valores reales:", rmse_porcentaje_sd, "%\n")
  # 
  # media_reales <- mean(y)
  # rmse_porcentaje_media <- (rmse_inferencia / media_reales) * 100
  # cat("RMSE (%) basado en la media de los valores reales:", rmse_porcentaje_media, "%\n")
  # 
  # cat('irep', irep, '\n')
  RMSE[irep] <- rmse_inferencia
  RMSERR[irep] <- rmse_inferencia_RR
  RMSERRCOMB[irep] <- rmse_inferencia_RRcom
  
  ## Krigeado
  
  ## Creo el objeto gstat
  s1 <- datostrain$s1
  s2 <- datostrain$s2
  coordinates(datostrain) <- ~s1 + s2
  # gkor <- gstat(formula = y ~ 1, data = datostrain)
      # gkun <- gstat(formula = y ~ s1 + s2, data = traint)
  gkun <- gstat(formula = y ~ s1 + s2 , data = datostrain)
      
  # emp_variogram_kor <- variogram(gkor)
  emp_variogram_kun <- variogram(gkun)
      
  # vgm_model_fit_kor <- autofitVariogram(
  #       formula = y~1,
  #       input_data = datostrain
  #     )
      
      # print(vgm_model_fit_kor$var_model)
      
      # vgm_model_fit_un <- autofitVariogram(
      #   formula = y ~ s1 + s2,
      #   input_data = traint
      # )
      
  vgm_model_fit_un <- autofitVariogram(
        formula = y ~ s1 + s2,
        input_data = datostrain
  )
      
      
      
      #print('hola')
      
  #Kr <- krigRes(datostrain, datostest, vgm_model_fit_kor$var_model, spam)[[1]]
      
      #print('adios')
      
  KrU <- krigRes(datostrain, datostest, vgm_model_fit_un$var_model, spam)  # [[2]]
  
  par(mfcol=c(1,2))
  
  par(mfrow=c(1,2))
  plot(pred_test_inferencia, test_inferencia$y)
  abline(a=0,b=1)
  plot(KrU$var1.pred, test_inferencia$y)
  abline(a=0,b=1)
  
  # RMSEKO[irep] = mean((datostest$y-Kr$var1.pred)**2)
  RMSEKU[irep] = sqrt(mean((datostest$y-KrU$var1.pred)**2))
  
  # modelo2_train=gam(formula=formula, data=train_inferencia)
  # predtot <- predict(modelo2_train, newdata = test_inferencia)
  # restot <- predtot - as.vector(test_inferencia$y)
  # rmse_tot <- sqrt(mean(restot^2))
  # RMSETOT[irep] <- rmse_tot
  
  
  
}

  
par(mfcol=c(1,1))  
boxplot(cbind(RMSE, RMSERRCOMB, RMSEKU), names = c('RGAM', 'GRGAM', 'UKr'), ylab='RMSE', outline=FALSE)
```


```{r}
plot(s2~s1, data=datos2,  type = "n")
points(s2~s1, data=train_inferencia, col=1,pch=16)
#points(s2~s1, data=test_inferencia, col=2,pch=16)
text(x = train_inferencia$s1,
     y = train_inferencia$s2,
     labels = row.names(train_inferencia), # Usa los nombres de las filas como etiquetas
     pos = 4, # Muestra el texto a la derecha del punto
     cex = 0.6, # Reduce un poco el tamaño del texto para que no se superpongan
     col = "blue")
```


